---
title: "Study of a possible association between level of education and job satisfaction"
date: June 2015
output:
  html_document:
    theme: cerulean
---
### Introduction:

Some people believe that a higher academic degree leads to a more satisfying work . A lively discussion is going on concerning this issue. This project puts this belief to a test by examining data from the General Social Survey (GSS). 

We try to explore a possible correlation between the respondent's highest educational degree and the level of satisfaction with the job he/she obtained later.

### Data:
The data represents a random sample of the US population. The cases are randomly selected individuals from the US population.

For the purposes of our study we will examine the following variables: **'satjob'** (On the whole, how satisfied are you with the work you do?) and **'degree'** (degree obtained by the respondents). Both variables are categorical and ordinal. The levels of 'satjob' can be ordered in the following decreasing order (Very Satisfied, Mod. satisfied, A Little Dissatisfied, Very Dissatisfied). The levels of 'degree' can be ordered in the increasing order as follows: Less Than High School, High School, Junior College, Bachelor, Graduate).

It is an observational study. Respondents were randomly selected from the US population. For this reason, the findings **can be generalized to the entire population**. Since this is an observational study and not an experiment (with random assignment to groups), **only a correlation** (association) can be established between the variables, **not a causal link**.

```{r}
load("statistics-project-gss.Rdata")
rawData <- nrow(subset(gss, !is.na(satjob)&!is.na(degree)))#40672
rawData <- cbind(gss["satjob"], gss["degree"])
```

### Exploratory data analysis:
- Number of rows with missing values (NA): **`r nrow(rawData)`**. We decided to drop NA values.
```{r}
data <- subset(rawData, !is.na(satjob) & !is.na(degree))
nrow(data)
```
Number of rows without missing values: **`r nrow(data)`** (rows lost: **`r nrow(rawData) - nrow(data)`**). Roughly 25% of data is lost. I will assume that NA values are randomly distributed and that no bias is introduced. It must be noted that it would be difficult to impute such a large number of missing values without introducing a bias. With this understanding, I decided to procede with a new data set which represents roughly 75% of the original data set and does not contain NA values.

- Contingency table - Total frequency per income
```{r results = "hide"}
library(gmodels)
joint <- CrossTable(data$satjob, data$degree)
```

```{r}
totalfreqSatjob <- joint$t
rowtotals <- as.vector(rowSums(totalfreqSatjob))
totalfreqSatjob <- cbind(totalfreqSatjob, rowtotals)
coltotals <- as.vector(colSums(totalfreqSatjob))
totalfreqSatjob <- rbind(totalfreqSatjob, coltotals)
totalfreqSatjob
```
The numbers of respondents belonging to each income group is not equal, hence it makes more sense to consider relative frequencies and not total counts.

- Contingency table - Relative frequency per income

```{r}
relfreqSatjob <- joint$prop.col
relfreqSatjob
```


- Joint count barplot

```{r fig.width = 4, fig.height= 4}
library(ggplot2)
ggplot(data=data, aes(x=degree)) + geom_bar(fill="#FF9999", colour="blue") + facet_wrap(~satjob)
```

- Joint relative frequency barplots per level of satisfaction

```{r}
df_rel = as.data.frame(relfreqSatjob)
colnames(df_rel) = c("Satisfaction", "Degree", "Frequency")
ggplot(data=df_rel, aes(x=Degree, y=Frequency)) + geom_bar(stat="identity", fill="#FF9999", colour="blue") + facet_wrap(~Satisfaction)+ theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



### Inference:
We have two categorical variables each of which has more than two levels. In this situation we use a **chi-squared independence test**.

- Null hypothesis -  H0 = job satisfaction and degree are independent
- Alternative hypothesis - Ha = job satisfaction and degree are dependent

- Conditions for chi-squared independence test
1. **Independence**. Sampled observations are independent since the sample is random, the sample represents less than 10% of the population and each case only contributes to one cell in the table.
2. **Sample size**. Each particular scenario (cell) has at least 5 expected cases. 

**First we will perform a chi-squared independence test manually and then apply the R function.**
- Compute the expected values:
```{r}
tbl <- table(data$satjob, data$degree); expected <- tbl
for (i in 1:4) {
  for (j in 1:5){expected[i,j] <- round(sum(tbl[i,]*sum(tbl[,j])/sum(tbl)), 0)}}
```

```{r echo = FALSE, results= "hide"}
rowSums(tbl)
rowSums(expected)
colSums(tbl)
colSums(expected)
```



```{r echo = FALSE}
expected
```
- Compute chi-squared:
```{r}
chi_sq <- sum((tbl-expected)^2/expected); chi_sq #tbl contains observed values
```
- Degrees of freedom: `(4-1)*(5-1)` = `r (4-1)*(5-1)`, p-value: 
```{r}
pchisq(chi_sq, 12, lower.tail = FALSE)
```
**P-value is close to zero, we reject the null hypothesis**

**Now we will compare this result with the result of the R's `chisq.test` function.**
```{r}
chisq.test(tbl)
```
p-value is far less than 0.05 at 95% significance level. **We reject the null hypothesis**.



### Conclusion:
We **reject the null hypothesis** that the level of job satisfaction and academic degree are independent and **accept the alternative hypothesis** - these variables are dependent. There might be a **positive correlation** between the level of degree and job satisfaction (i.e. respondents with a higher degree seem to be more satisfied with their work). Though, **we cannot conlcude that there is a casual link** since this is an observational study and not an experiment. It might be interesting to conduct a **further study** on other  possible factors contributing to the satisfaction at work (for example, level of happiness, optimism, pessimism, etc).

### References
- Data citation
Smith, Tom W., Michael Hout, and Peter V. Marsden. General Social Survey, 1972-2012 [Cumulative File]. ICPSR34802-v1. Storrs, CT: Roper Center for Public Opinion Research, University of Connecticut /Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributors], 2013-09-11. doi:10.3886/ICPSR34802.v1

- Persistent URL: http://doi.org/10.3886/ICPSR34802.v1

- A modified version of the survey (taken from another statistics class) was used for the purposes of this study.

```{r}
# library(ggplot2)
# counts <- ggplot(data=data, aes(x=degree, y=satjob)) + geom_bar(stat="identity")
# 
# counts_per_degree <- ggplot(data=data, aes(x=degree)) + geom_bar(fill="#FF9999", colour="blue") + facet_wrap(~satjob)
# 
# rel_very_satisfied <- ggplot(data, aes(x=degree, y=satjob))+ geom_bar(stat="identity",position='dodge')
# ggplot(data, aes(x=degree))+geom_density()
# 
# 
# ggplot(data, aes(satjob)) +
#     geom_bar(aes(y = ..count.. / sapply(PANEL, FUN=function(x) sum(count[PANEL == x])))) +
#     facet_wrap(~degree) 
# 
# df_rel = as.data.frame(relfreqSatjob)
# colnames(df_rel) = c("Satisfaction", "Degree", "Frequency")
# lt_hs_rel = subset(df_rel, Degree=="Lt High School")
# ggplot(data=lt_hs_rel, aes(x=Satisfaction, y=Frequency)) + geom_bar(stat="identity") #+ facet_wrap(~Satisfaction)
# ggplot(data=df_rel, aes(x=Satisfaction, y=Frequency)) + geom_bar(stat="identity") + facet_wrap(~Degree)
# ggplot(data=df_rel, aes(x=Degree, y=Frequency)) + geom_bar(stat="identity") + facet_wrap(~Satisfaction)
# 
# library(reshape)
# data.melt <- melt(data, id=c("degree"))
# ggplot(data.melt, aes(x=degree, y=value)) + geom_bar(stat="identity")
```

